Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-30 _tex/index.tex
Built site for gh-pages

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
could be generated. This study addresses these issues by representing
DCJ within the framework of causal inference. Specifically, utilizing
the structural approach, the study develops a scientific model to
clarify the causal assumptions and mechanisms inherent in the DCJ
system. It then translates this model into a probabilistic statistical
framework to estimate statistical relationships and infer causal effects
within the system. This research provides a robust probabilistic
foundation for the statistical analysis of DCJ data, building upon
Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
Its findings offer valuable insights for researchers and analysts
designing and implementing DCJ experiments.
\end{abstract}

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify the causal assumptions and
mechanisms inherent in the DCJ system. Next, using a minimal set of
assumptions, the study translates the scientific model into a
probabilistic statistical model. This model aims to produce statistical
estimates to draw inferences about plausible causal relationships within
the DCJ system.

@@

\section{Theoretical background}\label{sec-background}

\subsection{The structural approach to causal
inference}\label{sec-background-structural}

@@

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals form the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Despite this abstract nature, counterfactuals enable the
development of a \emph{theory of the world} that explains why specific
causes have specific effects and what occurs in their absence
\citep{Pearl_et_al_2018}. They achieve this by translating causal
statements into counterfactual statements, that is, statements about
``what would have happened in the world under different circumstances.''

@@

\subsection{DAGs, SCMs, and the flow of association and
causation}\label{sec-background-dag}

@@

\section{Theoretical framework}\label{sec-theory}

>>>>

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
could be generated. This study addresses these issues by representing
DCJ within the framework of causal inference. Specifically, utilizing
the structural approach, the study develops a scientific model to
clarify plausible causal assumptions and mechanisms inherent in the DCJ
system. The study then translates this model into a probabilistic
statistical framework to estimate statistical relationships and infer
causal effects within the system. This research provides a robust
probabilistic foundation for the statistical analysis of DCJ data,
building upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
researchers and analysts designing and implementing DCJ experiments.
\end{abstract}

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify plausible causal assumptions and
mechanisms inherent in the DCJ system. The study then translates the
scientific model into a probabilistic statistical model. This model aims
to produce statistical estimates to draw inferences about plausible
causal relationships within the DCJ system.

@@

\section{Theoretical framework}\label{sec-framework}

\subsection{The structural approach to causal
inference}\label{sec-framework-structural}

@@

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals form the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Nevertheless, despite their abstract nature,
counterfactuals enable the development of a \emph{theory of the world}
that explains why specific causes have specific effects and what occurs
in their absence \citep{Pearl_et_al_2018}. They achieve this by
translating causal statements into counterfactual statements, that is,
statements about ``what would have happened in the world under different
circumstances.''

@@

\subsection{Directed acyclic graphs (DAG)}\label{sec-framework-dag}

Graph theory is a branch of mathematics focused on the study of graphs.
Graphs are mathematical structures that model pairwise relations between
objects. They can represent physical relations, such as electrical
circuits and roadways, and less tangible structures, such as ecosystems
and sociological relations. Graphs have proven useful in various fields,
including computer science, operations research, and the natural and
social sciences \citep{Gross_et_al_2018}.

In statistics, one application that incorporates concepts from graph
theory is causal inference. Specifically, the structural approach to
causal inference uses directed acyclic graphs (DAG) to provide a formal
and graphical representation of a system's causal structure
\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates that the edges of the graph extend from a
\emph{parent} node to a \emph{child} node, with arrows showing the
direction of causal influence. The term \emph{acyclic} indicates that
the direction of causal influences does not form a loop, ensuring that
the influences do not cycle back on themselves \citep{McElreath_2020}.

A key advantage of DAGs is that, regardless of the complexity of the
causal structure, these structures can be composed using a minimal set
of building blocks. Moreover, representing any complex structure
requires only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These advantages imply that a
researcher seeking to depict and understand the causal assumptions of a
complex system only needs to decompose the structure into these basic
building blocks and then analyze the causal assumptions associated with
each block \citep{McElreath_2020}.

Figure~\ref{fig-dags} presents the five fundamental building blocks of
any DAG. Figure~\ref{fig-dag_bb1} illustrates two unconnected nodes,
representing an scenario where variables \(X_{1}\) and \(X_{2}\) are not
causally related. Figure~\ref{fig-dag_bb2} shows two connected nodes,
depicting an scenario where a \emph{parent} node \(X_{1}\) has a causal
influence on a \emph{child} node \(X_{2}\). This implies that variable
\(X_{2}\) is a \emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3}
illustrates a \emph{chain} (or \emph{pipe}), where \(X_{1}\) influences
\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). This DAG indicates that
\(X_{1}\) is a parent node of \(X_{2}\), and \(X_{2}\) is a parent node
\(X_{3}\). Furthermore, it shows that \(X_{1}\) is an \emph{ancestor} of
\(X_{3}\), with the relationship between the two variables being
mediated entirely by \(X_{2}\). Figure~\ref{fig-dag_bb4} depicts a
\emph{fork}, illustrating an scenario where variables \(X_{1}\) and
\(X_{3}\) are both influenced by \(X_{2}\). In this scenario \(X_{2}\)
is a parent node for both \(X_{1}\) and \(X_{3}\). Finally,
Figure~\ref{fig-dag_bb5} illustrates a \emph{collider} (also known as
\emph{inmorality}), where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\). Given the heuristic nature of DAGs, the use of fundamental
building block to construct a causal structure of a system is easier to
understand using a motivating example. The motivating example can also
serve to point out about other conventions when using DAGs.

\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb1.png}

\subcaption{\label{fig-dag_bb1}Two unconnected nodes}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb2.png}

\subcaption{\label{fig-dag_bb2}Two connected nodes or descendant}

\end{minipage}%
\newline
\begin{minipage}{0.33\linewidth}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb3.png}

\subcaption{\label{fig-dag_bb3}Chain or pipe}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb4.png}

\subcaption{\label{fig-dag_bb4}Fork}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb5.png}

\subcaption{\label{fig-dag_bb5}Collider or inmorality}

\end{minipage}%

\caption{\label{fig-dags}DAG's fundamental building blocks.}

Consider a system where variables \(X\) and \(Z\) influence a third
variable \(Y\). In this system, it is assumed that \(X\) and \(Z\) are
dependent on their own processes and are, therefore, independent from
each other. Figure~\ref{fig-dag_example1} presents the plausible causal
structure of this system. The DAG shows the endogenous variables
\(V=\{X,Z,Y\}\) as circled black nodes, indicating these variables are
observed. The arrows connecting the variables indicate the direction of
causal influence, while a lack of influence is indicated by the absence
of arrows. Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\)
represent everything else that is chosen not to be modeled explicitly.
These exogenous variables, or \emph{disturbances}, are usually
represented by open circles to indicate their unobserved nature.
Although this DAG explicitly shows the exogenous variables,
conventionally these are omitted for brevity, resulting in an equivalent
graph as shown in Figure~\ref{fig-dag_example2}.

\begin{minipage}{0.50\linewidth}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_example1.png}

\subcaption{\label{fig-dag_example1}Full DAG}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_example2.png}

\subcaption{\label{fig-dag_example2}Simplified DAG}

\end{minipage}%

\caption{\label{fig-example}DAGs for a plausible causal structure in a
system.}

\subsection{The flow of association and causation in
graphs}\label{sec-framework-flow}

@@

\section{Theory}\label{sec-theory}
